---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
___

## Loading and preprocessing the data

Before running analysis, ensure that your working directory is set to /RepData_PeerAssessment1 

Read in zipped csv file *activity.csv* into R

```{r}
activity <- read.csv(unz("activity.zip","activity.csv"),
                       colClasses = c("numeric","character","numeric")) 
```

___

## What is mean total number of steps taken per day?

####Create a table that reports the total steps per day

```{r}
total <- aggregate(steps ~ date,activity,sum)
total
```

####Create a histogram for total steps per day (total)
```{r}
library(lattice)
histogram(~ steps, total, type = "count") 
```

####Calculate and report the mean/median of total steps per day (total)
```{r}
mean(total$steps, na.rm = TRUE)
median(total$steps, na.rm = TRUE)
```

___

## What is the average daily activity pattern?

####Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
totalinterval <- aggregate(steps ~ interval,activity,mean)
xyplot(steps ~ interval, totalinterval, type = "l")
```

####Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
totalinterval[which.max(totalinterval$steps),1]
```

___

## Imputing missing values

####Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
colSums(is.na(activity))
```

####Devise a strategy for filling in all of the missing values in the dataset. 

Strategy = use the mean for of each 5-minute interval; if NA, then populate with that mean (rounded to nearest whole number)

####Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
#Run loop to populate NA with global mean for 5-minute interval (round to nearest whole number)
intervals <- unique(activity$interval)

for(i in 1:288) {
activity[activity$interval == intervals[[i]] & is.na(activity$steps),1] <- 
   totalinterval[totalinterval$interval == round(intervals[[i]],0),2]
}

#Check whether any NAs left unpopulated
colSums(is.na(activity))
```

####Make a histogram of the total number of steps taken each day  

```{r}
totalnona <- aggregate(steps ~ date,activity,sum)

histogram(~ steps, totalnona, type = "count")
```

####Calculate and report the mean and median total number of steps taken per day
```{r}
mean(totalnona$steps, na.rm = TRUE)
median(totalnona$steps, na.rm = TRUE)
```

####Do these values differ from the estimates from the first part of the assignment?

```{r}
summary(total)
summary(totalnona)
```

Yes - compare the summary statistics; the number of dates included in the analysis increases when populating the NAs and the overall distribution appears different when looking at summary quartile data.

####What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
plotold <- histogram(~ steps, total, type = "count",main = "Old (w/o NAs)")
plotnew <- histogram(~ steps, totalnona, type = "count", main = "New (w/ NAs)")

print(plotold, split = c(1,1,2,1), more = TRUE)
print(plotnew, split = c(2,1,2,1))
```

Compare the histograms for both distributions. The new data set with NAs populated is more concentrated at the mean than the data set where NAs where excluded.

___

## Are there differences in activity patterns between weekdays and weekends?

####Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 

```{r}
activity$date <- strptime(activity$date, format = "%Y-%m-%d")

activity$day <- weekdays(activity$date, abbreviate = TRUE)
activity$day <- factor(ifelse(activity$day == "Sat"|activity$day == "Sun","weekend","weekday"))
```

####Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval
(x-axis) and the average number of steps taken, averaged across all weekday days or weekend days
(y-axis).

```{r}
totalday <- aggregate(steps ~ as.character(date) + interval + day,activity,mean)
totaldayavg <- aggregate(steps ~ interval + day,totalday,mean)

xyplot(steps ~ interval | day, totaldayavg, type = "l", layout = c(1,2))
```